---
title: "DEG_TALS_nmtRNA"
output: html_document
date: "2025-04-10"
---

```{r}
setwd("/Users/shahada/Desktop/Research_Project/")
```




```{r}
library(ggplot2)
library(dplyr)
library(vsn)
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(RColorBrewer)
library(ComplexHeatmap)  # Load the ComplexHeatmap package
library(circlize)  # Load circlize for annotations
library(SummarizedExperiment)
library(tibble)
library(readxl)
```



#Load expression matrix and nuclear-encoded mitochondrial genes
```{r}
# Mitochondrial genes (nuclear-encoded)
mito_genes_TALS <- read_excel("Human.MitoCarta3.0 copy 2.xlsx", sheet = "A Human MitoCarta3.0")

# ALS-related genes (e.g., DEGs from brain bank)
als_genes_TALS <- read.delim("merged_cellular_TALS.txt")
```





#Filtering NEM genes from expression matrix
```{r}
als_mito_filtered_TALS <- als_genes_TALS[als_genes_TALS$gene %in% mito_genes_TALS$EnsemblGeneID_mapping_version_20200130, ]
```


```{r}
counts_nmtRNA_TALS <- als_mito_filtered_TALS
```


#Loading metadata
```{r}
coldata_nmtRNA_TALS <- read.delim("./phenotypeStarting_TALSmedial_SV1.txt", stringsAsFactors = FALSE, sep = " ")
```




#Converting Status column
```{r}
coldata_nmtRNA_TALS <- coldata_nmtRNA_TALS %>%
  mutate(Sex = recode(Sex, "1" = "Male", "2" = "Female"),
         Status = recode(Status, "1" = "Control", "2" = "Case"))
```




#Matching strings
```{r}
# Set the 'gene' column as rownames
rownames(counts_nmtRNA_TALS) <- counts_nmtRNA_TALS$gene
```




#Filter coldata from counts
```{r}
# Match samples between coldata and counts
samples_in_common <- intersect(rownames(coldata_nmtRNA_TALS), colnames(counts_nmtRNA_TALS))

# Subset counts matrix to only those samples (columns)
counts_nmtRNA_TALS <- counts_nmtRNA_TALS[, samples_in_common, drop = FALSE]
```


#Match samples, columns in expression matrix and rows in metadata
```{r}
all.equal(colnames(counts_nmtRNA_TALS), rownames(coldata_nmtRNA_TALS))
```



```{r}
se <- SummarizedExperiment(
    assays = list(counts_nmtRNA_TALS = as.matrix(counts_nmtRNA_TALS)),
    colData = coldata_nmtRNA_TALS)
```





#relevel Status factor to set "Control" as the reference group
```{r}
coldata_nmtRNA_TALS$Status <- relevel(factor(coldata_nmtRNA_TALS$Status), ref = "Control")
```






#Make sure all covariates are factors, create design formula for DESeq
```{r}
coldata_nmtRNA_TALS$Status <- as.factor(coldata_nmtRNA_TALS$Status)
coldata_nmtRNA_TALS$Sex <- as.factor(coldata_nmtRNA_TALS$Sex)
coldata_nmtRNA_TALS$AgeCat <- as.factor(coldata_nmtRNA_TALS$AgeCat)
coldata_nmtRNA_TALS$PMDCat <- as.factor(coldata_nmtRNA_TALS$PMDCat)
coldata_nmtRNA_TALS$RINCat <- as.factor(coldata_nmtRNA_TALS$RINCat)
dds_tals <- DESeqDataSetFromMatrix(countData = counts_nmtRNA_TALS, colData = coldata_nmtRNA_TALS, design = ~ Sex + AgeCat + PMDCat + RINCat + SV1 + Status, tidy = FALSE, ignoreRank = FALSE)
```





#Filter DEGs to include > 5 expression
```{r}
se <- dds_tals[rowSums(assay(dds_tals, "counts")) > 5, ]
nrow(se)
```




#run DESeq
```{r}
dds_tals <- DESeq(dds_tals)
```



#DE results
```{r}
results(dds_tals)
```



```{r}
res_TALS <- results(dds_tals)
```



```{r}
summary(res_TALS)
```



#Convert Ensembl gene identifiers to standardized gene symbols 
```{r}
if (!requireNamespace("biomaRt", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("biomaRt")

library(biomaRt)
```




```{r}
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```



```{r}
res <- results(dds_tals)
res_df <- as.data.frame(res)
res_df$ensembl_gene_id <- rownames(res_df)
```


```{r}
gene_annotations <- getBM(
  filters = "ensembl_gene_id",
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  values = rownames(res_df),
  mart = mart
)
```




```{r}
res_annotated_TALS <- merge(res_df, gene_annotations, by = "ensembl_gene_id", all.x = TRUE)
```




#Save output (Optional)
```{r}
write.csv(res_annotated_TALS, "DEGs_nmtRNA_TALS_all.csv")
```



#Filter for significant NEM DEGs
```{r}
all_deg_TALS <- res_annotated_TALS[which(res_annotated_TALS$padj < 0.05), ]
```



#Filter up and down regulated NEM DEGs
```{r}
up_deg_TALS  <- all_deg_TALS[which(all_deg_TALS$log2FoldChange > 0), ]
down_deg_TALS <- all_deg_TALS[which(all_deg_TALS$log2FoldChange < 0), ]
```




#Distribution of log2FC for significant genes
```{r}
# In R (assuming you used limma or DESeq2):
deg_sig_tals <- all_deg_TALS
hist(deg_sig_tals$log2FoldChange, breaks=100, main="Distribution of log2FC for significant genes")
```




#Calculate dispersion (biological variability)
```{r}
plotDispEsts(dds_tals)
```




#Volcano plot of sig NEM DEGs
```{r}
install.packages("ggrepel")  # only run once
library(ggrepel)
```



```{r}
library(ggplot2)

# Filter for NEM genes only
nem_deg_tals <- res_annotated_TALS

nem_deg_clean_tals <- nem_deg_tals[!is.na(nem_deg_tals$padj) & !is.na(nem_deg_tals$log2FoldChange), ]

top_genes_tals <- nem_deg_clean_tals[order(nem_deg_clean_tals$padj), ][1:20, ]


# Define significance threshold for adjusted p-value only
padj_thresh <- 0.05
nem_deg_clean_tals$significance <- ifelse(nem_deg_clean_tals$padj < padj_thresh, "Significant", "Not significant")


ggplot(nem_deg_clean_tals, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.8, size = 2) +
  geom_text_repel(data = top_genes_tals, aes(label = hgnc_symbol), size = 3, max.overlaps = 20) +
  scale_color_manual(values = c("gray", "red")) +
  theme_minimal() +
  labs(
    title = "TALS Significant NEM DEGs",
    x = "log2(Fold Change)",
    y = "-log10(adjusted p-value)",
    color = "Significance"
  )
```




#Over Representation Analysis (Gene Ontology (GO) and KEGG)
```{r}
install.packages("BiocManager")
BiocManager::install(c("clusterProfiler", "org.Hs.eg.db", "enrichplot", "ggplot2", "DOSE"))
```




```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
```




```{r}
library(dplyr)

# Replace `gene_id` with your actual column name if different
genes1_entrez <- bitr(all_deg_TALS$ensembl_gene_id, 
                      fromType = "ENSEMBL", 
                      toType = "ENTREZID", 
                      OrgDb = org.Hs.eg.db)
```

```{r}
deg1_merged <- inner_join(all_deg_TALS, genes1_entrez, by = c("ensembl_gene_id" = "ENSEMBL"))
```


#GO Biological Process
```{r}
go1 <- enrichGO(gene = deg1_merged$ENTREZID,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "BP",  # Options: "BP", "CC", "MF"
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.2)
```



#GO Cellular Component
```{r}
go_cc1 <- enrichGO(gene = deg1_merged$ENTREZID,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "CC",  # Changed from "BP" to "CC"
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.2)

```



#GO Molecular Function
```{r}
go_mf1 <- enrichGO(gene = deg1_merged$ENTREZID,
                OrgDb = org.Hs.eg.db,
                keyType = "ENTREZID",
                ont = "MF",  # Changed from "BP" to "CC"
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.2)
```




#KEGG annotation
```{r}
kegg1 <- enrichKEGG(gene = deg1_merged$ENTREZID,
                    organism = "hsa",  # Use "mmu" if mouse
                    pvalueCutoff = 0.05)
```





#ORA of NEM genes (plotting)
#Biological Process plot
```{r}
library(enrichplot)

# GO dotplot
dotplot(go1, showCategory = 10, title = "GO BP Enrichment - TALS")
```

#Cellular Component plot
```{r}
dotplot(go_cc1, showCategory = 10, title = "GO CC Enrichment - TALS")
```

#Molecular Function plot
```{r}
dotplot(go_mf1, showCategory = 6, title = "GO MF Enrichment - TALS")
```





#KEGG annotation
```{r}
# KEGG dotplot
dotplot(kegg1, showCategory = 10, title = "KEGG Enrichment - TALS")
```



















